---
title: 'Analysis with `RegEnrich`'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How can we use `RegEnrich` to identify key transcriptional regulators from RNA-seq data?
- What inputs does `RegEnrich` need (expression matrix, metadata, list of regulators)?
- Why do we need mouse-specific transcription factor (TF) information instead of the built-in human TFs?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Understand the overall purpose of `RegEnrich` in identifying key regulators (e.g. TFs).
- Load a mouse transcription factor list suitable for use with `RegEnrich`.
- Prepare an expression matrix, design matrix, and contrast for a `RegenrichSet` object.
- Run the main `RegEnrich` pipeline and inspect the resulting ranked regulators.

::::::::::::::::::::::::::::::::::::::::::::::::


```{r, message=FALSE,warning=FALSE,echo=FALSE}

#Libraries needed
library(edgeR)
library(goseq)
library(fgsea)
library(EGSEA)
library(clusterProfiler)
library(org.Mm.eg.db)
library(ggplot2)
library(enrichplot)
library(pathview)
library(edgeR)
library(impute)
library(preprocessCore)
library(RegEnrich)
library(STRINGdb)

#Load differential expression results

debasal <- read.csv("data/limma-voom_basalpregnant-basallactate", header = TRUE, sep = "\t")
deluminal <- read.csv("data/limma-voom_luminalpregnant-luminallactate", header = TRUE, sep = "\t")
seqdata <- read.csv("data/seqdata", header = TRUE, sep = "\t")

#Load Hallmark gene sets
load("data/mouse_hallmark_sets.RData")
# loads as Mm.H, each element = pathway

#Load metadata for samples
factordata <- read.table("data/factordata", header = TRUE, sep = "\t")

#Load filtered counts matrix 
filteredcounts <- read.csv("data/limma-voom_filtered_counts", header = TRUE, sep = "\t")


```

## Analysis with `RegEnrich`
`RegEnrich` is used to identify potential *key regulators* (e.g. transcription factors) that may be driving the gene expression changes observed in your RNA-seq experiment.

At a high level, the workflow looks like this:

- **Expression data**: log-transformed expression matrix (genes × samples).  
- **Differential expression**: identify genes that differ between groups (e.g. `limma`).  
- **Network construction**: build a regulator–target network (e.g. co-expression).  
- **Enrichment testing**: test whether targets of a regulator are enriched among DE genes.  
- **Ranking**: combine evidence to give each regulator a score and rank.


Before we set up `RegEnrich` properly, we will explore the default TF list that comes with the package and see why it is not appropriate for this mouse dataset.

:::::::::::::::::::::::::::::::::::: challenge

## Spot the problem: built-in TFs vs mouse data

1. Load the built-in transcription factor list:

   ```r
   data(TFs)
   ```
2. Inspect the TFs object:
- What kinds of identifiers are used (e.g. gene symbols, Entrez IDs)?
- Which species do these transcription factors belong to?

3. Based on what you see:
- Why might using TFs be a problem for our mouse expression dataset?
- What could go wrong in the analysis if we use human TFs with mouse RNA-seq data?
:::::::::::::::::::::::::::::::::::::::::::::::

## Using a mouse TF list from TcoF-DB
The TFs included in the package are human-only, so for mouse data we must provide our own list of mouse transcription factors.

For this workshop, we will use mouse TFs from [TcoF-DB](https://tools.sschmeier.com/tcof/home/).
You can directly download the file that we will be using from [this link](episodes/data/BrowseTF_TcoF-DB.csv).

The code below shows how to:

1. Load a mouse TF list from a CSV file.
2. Prepare an expression matrix for RegEnrich.
3. Create a RegenrichSet object.
4. Run the main RegEnrich pipeline and inspect the results.

```{r, eval=FALSE}
# Load mouse transcription factors (must include a "GeneID" column)
mouseTFs <- read.csv('data/BrowseTF_TcoF-DB.csv')

# Prepare expression matrix: genes x samples
logcounts <- filteredcounts[,4:15]
rownames(logcounts) <- filteredcounts$ENTREZID

# Convert to log CPM for RegEnrich
logcounts <- cpm(logcounts,log=TRUE)

# Define design (uses CellTyoeStatus metadata) and example contrast
design = model.matrix(~ factordata$CellTypeStatus)
contrast = c(-1, 1,0,0,0,0) 

# Initialise a RegenrichSet object
object = RegenrichSet(expr = logcounts,
                      colData = factordata,
                      reg = unique(mouseTFs$GeneID), # regulators
                      method = "limma", # differential expression analysis method
                      design = design, # design model matrix
                      contrast = contrast, # contrast
                      networkConstruction = "COEN", # network inference method
                      enrichTest = "FET") # enrichment analysis method

print(object)





```

:::: caution

The `regenrich_diffExp` step can take a while. We have already run this step for you and you can [download the `object` data directly using this link](episodes/data/regenrich_object.RData).


:::


```{r, eval=FALSE}

# Perform RegEnrich analysis
set.seed(123)

# This step takes a while
object = regenrich_diffExpr(object) %>%
  regenrich_network() %>%
  regenrich_enrich() %>%
  regenrich_rankScore()


# Obtain results (ranked regulators)
res = results_score(object)
print(res)

# Visualise regulator-target expression for selected regulator
plotRegTarExpr(object, reg = "71371")


```




```{r, echo=FALSE,warning=FALSE}
# Load mouse transcription factors (must include a "GeneID" column)
mouseTFs <- read.csv('data/BrowseTF_TcoF-DB.csv')

# Prepare expression matrix: genes x samples
logcounts <- filteredcounts[,4:15]
rownames(logcounts) <- filteredcounts$ENTREZID

# Convert to log CPM for RegEnrich
logcounts <- cpm(logcounts,log=TRUE)

# Define design (uses CellTyoeStatus metadata) and example contrast
design = model.matrix(~ factordata$CellTypeStatus)
contrast = c(-1, 1,0,0,0,0) 

# Load object instead
load('data/regenrich_object.RData')

# # Initialise a RegenrichSet object
# object = RegenrichSet(expr = logcounts,
#                       colData = factordata,
#                       reg = unique(mouseTFs$GeneID), # regulators
#                       method = "limma", # differential expression analysis method
#                       design = design, # design model matrix
#                       contrast = contrast, # contrast
#                       networkConstruction = "COEN", # network inference method
#                       enrichTest = "FET") # enrichment analysis method
# 
# print(object)
# 
# 
# # Perform RegEnrich analysis
# set.seed(123)
# 
# # This step takes a while
# object = regenrich_diffExpr(object) %>%
#   regenrich_network() %>%
#   regenrich_enrich() %>%
#   regenrich_rankScore()

# Obtain results (ranked regulators)
res = results_score(object)
print(res)

# Visualise regulator-target expression for selected regulator
plotRegTarExpr(object, reg = "71371")


```

:::: spoiler

## Understanding design matrices and contrasts

`RegEnrich` uses a design matrix and contrast in a similar way to limma: they define which groups you want to compare.

We create a design matrix from a factor in our sample metadata:

`design <- model.matrix(~ factordata$CellTypeStatus)`


This turns the factor `CellTypeStatus` into one column per group (plus an intercept).
A contrast vector then specifies how to combine these columns to define a comparison.

For example, a contrast like:

`contrast <- c(-1, 1, 0, 0, 0, 0)`

means:

- Compare group 2 vs group 1
- i.e. “group 2 MINUS group 1”
- All other groups are ignored (set to 0)

The exact mapping of positions in the contrast to group names depends on the order of the factor levels in `factordata$CellTypeStatus`.


::::





:::::::::::::::::::::::::::::::::::: challenge

## Test your understanding: contrasts

Look at the factor levels in `factordata$CellTypeStatus`:

```r
levels(factordata$CellTypeStatus)
```
1. How many groups are there?
2. Which group is used as the baseline (reference) in the design matrix?
3. Write a contrast that compares Luminal pregnant vs Basal pregnant.
4. In words, what biological question does that contrast represent?

:::::::::::::::::::::::::::::::::::::::: solution

The number of groups equals the number of unique levels returned by
`levels(factordata$CellTypeStatus)`

The baseline group is the first level of the factor.

If the factor levels are ordered like:

[1] "Basal pregnant" "Basal lactate" "Luminal pregnant"
    "Luminal lactate" "Stem" "Other"


Then the corresponding contrast to compare
Luminal pregnant vs Basal pregnant is:

`contrast <- c(-1, 0, 1, 0, 0, 0)`


This means:

1 → Luminal pregnant

-1 → Basal pregnant

0 → all other groups ignored

The biological question this is answering is:

“Which transcriptional regulators differ between Luminal pregnant and Basal pregnant samples?”

That is, regulators that functionally distinguish these two cell states.

:::::::::::::::::::::::::::::::::::::::::::::::
:::::::::::::::::::::::::::::::::::::::::::::::


## Inspecting and interpreting RegEnrich results
The `results_score(object)` call returns a table of regulators with associated statistics.
Typical columns summarise:
- The regulator identifier (e.g. Entrez ID or gene symbol)
- Evidence from differential expression and/or network structure
- A combined score used to rank regulators (higher often = more influential)

A simple way to start exploring is to look at the top regulators and their expression patterns across conditions:
- Are top-ranked regulators differentially expressed between groups?
- Do their predicted targets show coordinated expression changes?
- Does the expression of a regulator and its targets match your biological expectations?

:::::::::::::::::::::::::::::::::::: challenge
## Interpreting regulator results
Using the output table `res`:
- Identify the top 3 regulators by whatever ranking column is provided (e.g. `rankScore`).
- For one of these regulators, check its expression across samples using `plotRegTarExpr()`.
- Does this pattern support the idea that this regulator is involved in the contrast you specified?
- How might you follow this up experimentally?
:::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints 

- `RegEnrich` helps identify potential regulatory drivers (e.g. TFs) behind observed gene expression changes.
- The package’s built-in TF dataset `(data(TFs))` is human-specific and not suitable for mouse RNA-seq analysis.
- For mouse data, a mouse-specific TF list (e.g. from TcoF-DB) must be supplied via the reg argument.
- A RegenrichSet object requires: an expression matrix, sample metadata, a regulator list, and a design/contrast specification.

::::::::::::::::::::::::::::::::::::::::::::::::

